Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-28 wpe.tex
more work on the wpe section

For the purposes of this paper, we view the Shannon entropy---in
particular its growth rate with respect to word length (the
\emph{Shannon entropy rate})---as a measure of complexity and
unpredictability in a time series.  A time series with a high entropy
rate is almost completely unpredictable; conversely one with low
entropy rate is often quite predictable. {\color{red} [[does the claim
      in the previous sentence need a citation, a justification, or an
      ``in practice,''?]]}  This can be made more rigorous: Pesin's
relation \cite{pesin77} states that in chaotic dynamical systems, the
Kolmogorov-Sinai (KS) entropy is equal to the sum of the positive
Lyapunov exponents, $\lambda_i$.  The Lyapunov exponents directly
quantify the rate at which nearby states of the system diverge with
time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta
x(0) \right|$.  The faster the divergence, the larger the entropy, and
the more difficult prediction becomes.  The KS entropy is defined as
the supremum of the Shannon entropy rates of all
partitions~\cite{petersen1989}. The partition that achieves this
supremum is the \emph{generating partition} discussed in
Section~\ref{sec:related}.

@@

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.
With this in mind, it is common in the literature to normalize
permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With
this convention, ``low'' PE is close to 0 and ``high'' PE is close to
1. Finally, it should be noted that the permutation entropy has been
shown to be identical to the Kolmolgorov-Sinai entropy for many large
classes of systems \cite{amigo2012permutation}, so long as
observational noise is sufficiently small. As mentioned before, this
is equal to the Shannon entropy rate of a generating partition of the
system.  {\color{red} [[What this normalization allows us to do, then,
      is approximate the redundancy of a signal by $1 -
      \frac{H(\ell)}{\log_2(\ell!)}$.]] [[This sentence is out of
      place.  Shouldn't it be before the sentence that starts with
      ``Finally...''?]]}

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise.  {\color{red} The weighted permutation
  entropy would be dominated by the switching rather than by the
  stochastic fluctuation. To accomplish this, [[I am completely
      confused by the wording here.  Do you mean that the {\bf
        original} PE would be dominated?  And that you don't want
      that?  If not, you need to fix this wording -- i.e., ``would''
      to ``should'' and an explanation of why one would want to
      calculate a PE that captures the switching.]]} the \emph{weight}
of a permutation is taken into account:
\begin{align*}
  w(x_{t+1:t+\ell}) = \frac{1}{\ell} \sum_{x_i \in \{x_{t+1:t+\ell}\}}
                      \left( x_i - \bar{x}_{t+1:t+\ell} \right)^2
\end{align*}
where $x_{t+1:t+\ell}$ is a sequence of values $x_{t+1}, \ldots,
x_{t+\ell}$, and $\bar{x}_{t+1:t+\ell}$ is the arithmetic mean of
those values.

The weighted probability of a permutation is then:
\begin{align*}
  P_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell}) \cdot \delta(\phi(x_{t:t+\ell}), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell})}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively,
this weighted probability enhances permutations that are involved in
``large'' features and de-emphasizes permutations that are small in
amplitude relative to the features of the time series.  {\color{red}
  [[The weighted permutation entropy is then:]] [[you used exactly
      these words above.  what is the textual causality of this
      ``then'' and that ``then''?]]}
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P_w(\pi) \log_2 P_w(\pi),
\end{align*}
which can also be normalized {\color{red} [[why?]]}  by dividing by
$\log_2(\ell!)$.  This normalization is used in all the results that
follow.

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration here is that the value be large enough that
forbidden ordinals are discovered, yet small enough that reasonable
statistics over the {\color{red} [[existing??]]}  ordinals are
gathered.  If an average of 100 counts per ordinal is considered to be
sufficient, for instance, then $\ell =
\operatornamewithlimits{argmax}_n \{ T \gtrapprox 100 n! \}$.  In the
literature, $3 \le \ell \le 6$ is a standard choice---generally
without any formal justification.  In theory, the permutation entropy
should reach an asymptote with increasing $\ell$, but that can require
an arbitrarily long time series. In practice, what one should do is
calculate the \emph{persistent} permutation entropy by increasing
$\ell$ until the result converges, but data length issues can intrude
before that convergence is reached.  We used this approach to choose
$\ell = 6$ for the experiments in the following section.  This value
represents a good balance between accurate ordinal statistics and
finite-data effects.

>>>>

For the purposes of this paper, we view the Shannon entropy---in
particular its growth rate with respect to word length (the
\emph{Shannon entropy rate})---as a measure of complexity and
unpredictability in a time series.  Time-series consisting of IID 
random variables, such as white noise, have maximal entropy rates, 
whereas highly structured time-series, for example periodic, have very
low or zero entropy rate. A time series with a high entropy rate is 
almost completely unpredictable; conversely one with low entropy rate
is often quite predictable. This can be made more rigorous: Pesin's
relation \cite{pesin77} states that in chaotic dynamical systems, the
Kolmogorov-Sinai (KS) entropy is equal to the sum of the positive
Lyapunov exponents, $\lambda_i$.  The Lyapunov exponents directly
quantify the rate at which nearby states of the system diverge with
time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta
x(0) \right|$.  The faster the divergence, the larger the entropy, and
the more difficult prediction becomes.  The KS entropy is defined as
the supremum of the Shannon entropy rates of all
partitions~\cite{petersen1989}. The partition that achieves this
supremum is the \emph{generating partition} discussed in
Section~\ref{sec:related}.

@@

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.
With this in mind, it is common in the literature to normalize
permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With
this convention, ``low'' PE is close to 0 and ``high'' PE is close to
1. Finally, it should be noted that the permutation entropy has been
shown to be identical to the Kolmolgorov-Sinai entropy for many large
classes of systems \cite{amigo2012permutation}, so long as
observational noise is sufficiently small. As mentioned before, this
is equal to the Shannon entropy rate of a generating partition of the
system. This transitive chain of equalities, from permutation entropy 
to Shannon entropy rate via the KS entropy, then allows us to
approximate the redundancy --- being the dual of the Shannon entropy 
rate --- of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE is dominated by the noise about 
the fixed points driving it to be $\approx 1$, in some sense hiding
the fact that the signal is actually quite structed. The weighted 
permutation entropy on the other hand would be dominated by the 
switching rather than by the stochastic fluctuations. To accomplish 
this, the \emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{t+1}^\ell) = \frac{1}{\ell} \sum_{i = t}^{t+\ell}
                      \left( x_i - \bar{x}_{t+1}^\ell \right)^2
\end{align*}
where $x_{t+1}^\ell$ is a sequence of values $x_{t+1}, \ldots,
x_{t+\ell}$, and $\bar{x}_{t+1}^\ell$ is the arithmetic mean of
those values.

The weighted probability of a permutation is defined as:
\begin{align*}
  P_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1}^\ell) \cdot \delta(\phi(x_{t+1}^\ell), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1}^\ell)}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively,
this weighted probability enhances permutations that are involved in
``large'' features and de-emphasizes permutations that are small in
amplitude relative to the features of the time series. Using the
standard form of an entropy, the weighted permutation entropy is then:
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P_w(\pi) \log_2 P_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(\ell!)$, making $0
\le \textrm{WPE} \le 1$.  This normalization is used in all the
results that follow.

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration here is that the value be large enough that
forbidden ordinals are discovered, yet small enough that reasonable
statistics over the ordinals are
gathered.  If an average of 100 counts per ordinal is considered to be
sufficient, for instance, then $\ell =
\operatornamewithlimits{argmax}_n \{ T \gtrapprox 100 n! \}$.  In the
literature, $3 \le \ell \le 6$ is a standard choice---generally
without any formal justification.  In theory, the permutation entropy
should reach an asymptote with increasing $\ell$, but that can require
an arbitrarily long time series. In practice, what one should do is
calculate the \emph{persistent} permutation entropy by increasing
$\ell$ until the result converges, but data length issues can intrude
before that convergence is reached.  We used this approach to choose
$\ell = 6$ for the experiments in the following section.  This value
represents a good balance between accurate ordinal statistics and
finite-data effects.

As a final note: there has been prior work under a similar title to
ours~\cite{haven2005}, but there are only superficial similarities 
between the two research projects.  Haven {\sl et al.} utilize the 
relative entropy to quantify how much better a distribution over 
system states evolved from small ensembles predicts the future of a 
dynamical system compared to its asymptotic distribution. Our work 
quantifies the predictability of a single observed time series using 
weighted permutation entropy and makes no assumptions about the 
underlying dynamics. These approaches are very different: Haven's 
compares two distributions, the average behavior versus the behaviors 
that can be seen given a reasonably known state in the past; ours 
attempts to quantify how well an appropriate predictor will do for a 
given time-series without any understanding of how that time-series
was generated.
